Predicting the conformational preferences of proteins using a physics-based free energy method

### 2013-05-12 Paper/paper.tex
Further tweaks.

The calculation of free energy differences is of central importance in the simulation of biochemical
systems. It is particularly difficult to calculate between pairs of macromolecular conformations as
well as a computationally expensive task with existing methods. In this work, the confinement approach
is used to calculate absolute free energies of biomolecular systems. This method provides two main
advantages: it does not require a reaction coordinate or transition path and it is fast to compute.
The calculated free energy can be decomposed into a per residue contribution in an approximate way. Per
residue free energy allows us to identify the reason behind conformational preferences in
biomolecules. Through out the article we show its use in different challenging modeling problems. In
particular, we show its use in predicting the conformational preference of chamaleon sequences
(sequences with high sequence identity and different folds). This sequence dependent conformational
preferences and per residue free energy decomposition set the stage for the use of this method in
protein design.

@@

In some problems of protein science, you want to know the relative stability of a protein's conformation $A$ compared to
its conformation $B$.  We call this the \emph{difference free energy}. For example, in all allosteric mechanisms, a protein
often adopts one conformation when a ligand is bound and another conformation when no ligand is bound~\cite{Elber2007}.
To understand the mechanism requires knowledge of the relative free energies of the protein conformations.  Another
example is in protein structure prediction, for example as practiced in the Critical Assessment of Structure Prediction
(CASP) competition~\cite{Moult2011}. If you are using some computational model and have predicted two putative native
structures, $A$ and $B$, you want to compute which has the lower free energy, in order to know which is the more
native-like.  In a third example, you may want to know whether mutating a few amino acids in a protein could cause the
protein to switch from one stable conformation to another, because that could have important consequences for biological
mechanisms and disease. And fourth, sometimes binding a ligand induces a protein from one conformation to another. A
quantitative understanding of such \emph{induced-fit} situations requires knowledge of the difference free energy of the
protein in the two states. In all these cases, it would be useful to have a computational method that can be used
efficiently with atomically detailed physical forcefield models to compute the free energy differences between different
given protein conformations.

One widely explored strategy
is to use molecular dynamics simulations along some putative reaction coordinate pathway from conformation $A$ to
$B$~\cite{Elber2007,West2007,Haas2007,Jonsson1998,E2007,Dellago2002,Cheng2006,Elber2005,Chipot2007}. The free energy along this
reaction coordinate can then be determined using methods such as umbrella sampling \cite{Torrie1977,Mascarenhas2013}  
and weighted histogram analysis method (WHAM) \cite{Kumar1992}. Such
approaches have several limitations. First, it is necessary to know an efficient reaction pathway from $A$ to $B$. If
conformations $A$ and $B$ are quite different, then it can be challenging to find such paths. Second, these methods are
computationally slow. To get an accurate estimate of the total free energy difference $\Delta G = G_B - G_A$ requires
accurate determinations of the many small free energy differences $A \rightarrow 1 \rightarrow 2 \ldots \rightarrow B$,
so each step requires substantial amounts of sampling.  And third, these methods are prone to large errors, because the
pathway error is a sum of many errors among the many steps.

The calculation of protein conformational free energy has been successfully attempted by a number of groups
\cite{Ytreber2006a,Shell2010,Ytreberg2006,Zheng2008,Spichty2010,Strajbl2000,Park2008,Tyka2006,Cecchini2009,Ovchinnikov2013}. Some of
these methods like the reference system method~\cite{Ytreberg2006}, deactivated morphing~\cite{Park2008}, the
confinement method~\cite{Tyka2006,Cecchini2009,Ovchinnikov2013} take an alternative strategy for computing difference
free energies that does not require knowing a pathway from $A$ to $B$.

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=3.5 in]{method.pdf}
    \end{center}
    \caption{Graphical representation of the thermodynamic cycle involving confinement method.}
\label{fig:method}
\end{figure}

@@

First, we validated that our implementation of the confinement method produces results compatible with previous
calculations reported in the literature.  The method has previously been applied to a 16 amino acid residue
$\beta$-hairpin from protein G, known as BHP~\cite{Cecchini2009}. We calculated the free energy difference between the
native conformation---called bhp1, which has a two-stranded $\beta$-sheet---and a non-native conformation---called bhp3,
which has a three-stranded $\beta$-sheet. Our confinement calculation shows that bhp1 is more stable by 1.7 kcal/mol,
consistent with 4 $\mu$s equilibrium simulations showing that bhp1 is more favorable configuration by 1.8 kcal/mol, and
in agreement with previous calculations~\cite{Cecchini2009}.

Second, we looked at 6 target proteins from the CASP9 experiment~\cite{Moult2011}.  For each target, we examined up to 5 submitted
models.  We computed the difference free energy between the true native and the best model.
Figure~\ref{fig:summary_casp} and Table~\ref{table:casp_control} show that, in 5 out of 6 cases, the confinement method
assigns a lower free energy to the experimentally determined structure than to any of the decoys.  Other well-known
discriminators can also successfully tell native structures from computer generated models~\cite{Sheffler2009,Zhou2002} 
; it is just an independent useful validation that the confinement calculations make
sense.

\begin{table}
\begin{center}
\caption{The confinement method assigns a more favorable free energy to the experimentally determined structure than to
    computer-generated predictions. For each target, we examined as many as five predictions submitted by CASP
    participants. We report the free energy difference between the most favorable decoy and the experimentally
    determined structure. Positive $\Delta\Delta G (=G_{best~decoy} - G_{native})$ values indicate that the experimental structure is predicted to be
    more favorable than any of the decoys.}
\label{table:casp_control}
\begin{tabular}{l l l l}\hline
    CASP Target  & PDB Identifier & $\Delta \Delta G$ (kcal/mol) & GDT-TS \\ \hline
     T0531       &    2KJX        &          $11.15 \pm 0.70$    &  $44$ \\ \hline
     T0538       &    2L09        &          $-3.00 \pm 0.47$    &  $96$ \\ \hline
     T0540       &    3MX7        &          $16.94 \pm 0.49$    &  $70$ \\ \hline
     T0559       &    2L01        &          $2.10 \pm 0.24$     &  $94$ \\ \hline
     T0560       &    2L02        &          $18.00 \pm 0.49$    &  $94$ \\ \hline
     T0569       &    2KYW        &          $20.01 \pm 0.69$    &  $78$ \\ \hline
\end{tabular}
\end{center}
\end{table}
\Alberto{Should we comment on T0560? Why is it giving such a high value with such high gdt? It's not
the point to make here, but it is what jumps at me.}
\Arijit{I think we have made it clear that crystal structures are often more stable due to salt-bridge effects. Also, we have shown
in other cases that sidechains orientation may play important roles inspite of close GDT-TS. You will see free energy difference between 
two computer generated model is not so high. So, I believe sidechain may play some role here.}

@@

We also tested the confinement method predictions for difference free energies on the chameleon sequences of Orban et
al.\cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012}. These are instances in which two highly
similar sequences fold into remarkably different structures. Orban and co-workers have designed a protein-G-like
sequence of 56-residues that is marginally stable in one of two possible folds. By mutating key residues in this
sequence they are able to stabilize one fold or the other (see Figure~\ref{fig:orban}). We refer to the $4\beta +
\alpha$ structure as the $\beta$ conformation, and we refer to the $3\alpha$ structure as the
$\alpha$ conformation. We denote sequences that prefer the $\alpha$ fold as GA and sequences that
prefer $\beta$ as GB. One
pair of sequences (GA88/GB88) is 88 percent identical in sequence, differing in seven positions. Another pair
(GA95/GB95) is 95 percent identical, differing in three positions. Accurately predicting the structural preferences of
these structures presents a serious challenge for computational
methods \cite{Allison2011}.

@@

We initially approached this problem by making a model of each sequence with the same backbone structure as its partner
chameleon sequence. For example, we took the sequence of GA88 and built a model with the same overall structure as GB88.
We then used the confinement method to assess the free energy difference between the experimentally determined structure
of GA88 and the model (with the GA88 sequence and the GB88 structure). The confinement method was able to predict the
conformational preferences correctly for all four sequences (data not shown). This is, however, not surprising as it is
often easy to distinguish computational models from native structures~\cite{Handl2009,Sheffler2009}. To avoid this
potential problem, we instead computed relative free energy of two different computer generated models for each
sequence. One model is based on the $\alpha$ structure and the other on the $\beta$ structure (see Supporting
Information for details on the modeling procedure). This is a much more realistic test of the confinement method's
ability to accurately calculate the difference free energy.

@@

To better understand the mechanism behind these chameleon proteins, we decomposed the calculated free energy into
per-residue contributions in an approximate way~\cite{Tyka2006}. We decompose each confinement step ($\Delta
G_{A,A^\ast}$, $\Delta G_{B,B^\ast}$) into per-residue contributions. We can also decompose the remaining enthalpy in
the confined ensemble into per-residue contributions. However, the method is approximate because we do not include the
residual conformational entropy from the normal mode or quasiharmonic analysis steps. Computing per-residue
contributions helps us to identify important residues that stabilize a particular conformation. The per residue free
energy, $\Delta \Delta G (\beta - \alpha)$ is shown in Figure~\ref{fig:perresidue_orban}.

@@

Although the overall free energy difference between the two structures is small (within 3--5 kcal/mol), individual
residues can show marked preferences for being in either $\alpha$ or $\beta$ conformation. Such differences can be
understood by looking in detail at the local environment for those residues. For example, the region around residue 7
forms a random coil in the $\alpha$ structure, whereas it is forming a beta sheet structure in the $\beta$ structure.
These residues strongly favour the locally well packed and hydrogen bonded environment found in the $\beta$ sheet.
Overall, favoring $\alpha$ or $\beta$ structure is a delicate balance, where the relative global free energy difference
is small and the contributions of different per residue tendencies balance out. It is therefore very likely that by
changing key residue the global fold preferences can be changed.

Three residues are mutated between GA95 and GB95, at positions 20, 30, and 45. In GA95, these residues (L20, I30 and
L45) stabilize the $\alpha$ structure (compare the upper and lower panels on Figure~\ref{fig:perresidue_orban}). In
GB95, two of these residues (F30 and Y45) favor the $\beta$ structure, because they have large solvent exposed surface
areas in the $\alpha$ structure, but are more buried in the $\beta$ structure. Additionally Y45 forms a hydrogen bond
with D47 in the $\beta$ structure. On the other hand, residue A20 from GB95 still favors the $\alpha$ structure,
although less strongly than L20 in GA95.

The $\alpha$ and $\beta$ sequences are nearly identical, so there are some common features observed for all sequences.
The experimental observations classified the protein into two parts: Amino acids 9--51 are fully structured in both
folds, whereas residues 1--8 and 52--56 are unstructured in the $\alpha$ fold, but form $\beta$-strands in the $\beta$
fold. Most of the amino acid residues in the region 1--9 have negative per-residue free energies, which means that these
residues favor the $\beta$ structure. Roles of some other important residues in stabilizing either the $\alpha$ or
$\beta$ conformation are summarized in Figure S1.

@@

These per-residue free energy decompositions provide a great deal of insight into the driving forces behind protein
folding and conformational change. We believe that such calculations may also be useful for protein design---designing
proteins with specific structures and functions.

@@

We have performed two experiments centered on CASP. In the first experiment, we tried to rank-order predictions for
several targets. For each target, the predictions were either produced by a single group---presumably using the same
method for each prediction, or were produced by several different groups---using different methods. The goal of this
experiment is to determine if the physics-based confinement method can correctly identify more native-like structures as
having lower free energies. Our second experiment was to see if the confinement method can identify structures that are
missed by other meta-prediction servers. Most successful meta-prediction servers are based on the idea of consensus: if
many different prediction methods produce similar results, then that is probably a correct prediction \cite{Kryshtafovych2011,Wang2011}. This is
often a powerful heuristic, but it can miss cases where there is a very good result that is only predicted by one
method. We chose several cases where such structures were missed by the best meta-predictors in CASP and assessed if the
confinement method can correctly identify these accurate models.

As is common in the CASP experiment, we assess our results in terms of Global Distance Test Total Score (GDT-TS)
\cite{Zemla2003}, which is a C$\alpha$ based measure of structural accuracy. It can be understood roughly as the
percentage of residues that are correctly positioned in the model (range 0 to 100, higher is
better). We did not have enough
computer power to analyze every model, so the initial models for the test were chosen from a selection of different
server groups that have done well in past CASP events.

@@

\begin{table}
\begin{center}
\caption{The six different CASP targets that are used in this study. The corresponding PDB Identifier and the 
         organism type are also listed.}
\label{table:casp_testcase}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier &  Organism type  \\ \hline
     T0531       &    2KJX        &  extracellular domain of the jumping  \\
                 &                &  translocation breakpoint protein  \\ \hline
     T0538       &    2L09        &  protein asr4154 from Nostoc sp. PCC7120     \\ \hline
     T0540       &    3MX7        &  human Fas apoptotic inhibitory molecule     \\ \hline
     T0559       &    2L01        &  protein BVU3908 from Bacteroides vulgatus   \\ \hline
     T0560       &    2L02        &  protein BT2368 from Bacteroides  \\ 
                 &                &  thetaiotaomicron         \\ \hline
     T0569       &    2KYW        &  domain of adhesion exoprotein from \\ 
                 &                &   Pediococcus pentosaceus \\ \hline
\end{tabular}
\end{center}
\end{table}

@@

We performed a similar calculation for target T0560 with two models from the group ``Splicer''. The remaining three
models were discarded as they were too similar to the rest of the models. Again, we can correctly identify the native
sate and our calculated free energy based ranking matches well with GDT-TS (Figures S2 in the supporting information).

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=3.5 in]{T0540.pdf}
    \end{center}
    \caption{The confinement method can correctly rank the native structure and two models submitted by different
        prediction groups.}
\label{fig:T0540}
\end{figure}
\Alberto{if the drop down bars/point are a different color it will be easier to see when the crystal
is on top of the axis}
\Arijit{I will wait for Sarina's figure. I have changed the caption of figure 5 to say why some region are colored brown.}

\subsection*{The per-residue free energy is sensitive to small changes in protein conformation:}

In a previous section, we discussed how the per-residue free energy can reveal mechanistic
detail behind the conformational preference of a chameleon sequence. In this section, our aim is to apply the same
method and try to understand if the per-residue free energy can help us identify residues which stabilize or destabilize
a particular region of a protein. In this case, the two conformations have similar folds with only small changes in
localized regions. We chose CASP target T0569 and compared the experimental NMR structure with the best predicted model
(GDT-TS=78; predicted by the ``Mufold'' group). The confinement method predicts that the experimental structure is more stable
by 20 kcal/mol (Figure S2).

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=3.5in]{T0569_perres1.pdf}
    \end{center}
    \caption{Per-residue difference free energy between the experimental NMR structure and the best prediction for CASP
        target T0569. The amino acid residues that are colored in deep red and deep blue stabilizes the NMR structure
        and the prediction, respectively; the residues with light blue color does not have a strong preference.}
\label{fig:T0569_per_residue}
\end{figure}

\subsection*{The confinement method occasionally produces incorrect results:}

@@

\begin{figure}
\begin{center}
\includegraphics[width=4.9 in,height=4.0 in]{Target_538_compare.pdf}
\end{center}
\caption{The per residue free energy calculation found some differences of the side chain orientation in computer generated model 1 and 
experimental structure of CASP target T0538. 
(A) Two salt bridges between residues E35-R32 and E28-K24 exist in the native structure. These are
absent in the computer generated model structure. Instead
(B) a new salt bridge is formed between residues R32-E28 in model. (C) Another salt bridge between
residues  R26-E50 in the experimental structure. (D) In comparison to the native 
structure, orientation of F51 differs in the model.}
\label{fig:T0538compare}
\end{figure}

@@

A part of the CASP experiment is dedicated to the quality assessment (QA) of predicted models \cite{Kryshtafovych2011}.
Here, predictors were asked to score each model on a scale (known as qmode) from 0 to 1, with higher values
corresponding to better models \cite{Kryshtafovych2011}. Most of the groups use consensus strategies in such experiments.
Here, we chose two computer generated models from CASP taget T0538, where the top performing group "MUFOLD-WQA" \cite{Wang2011} failed.
One of the model was from "PconsR" (GDT-TS = 83, qmode 1 = 0.5865) and another from "MULTICOM-NOVEL" ( GDT-TS = 83, qmode 1 = 0.5865). We 
choose these two models as the model by PconsR was the most accurate
predicted model for this target, whereas the other model was predicted best by QA test of MUFOLD-WQA. The calculated
free energy using confinement method indicate that the model by PconsR is more stable by $3.9
Kcal/Mol$ which supports
the GDT\_TS trend.  Although consensus prediction strategies do well specially when dealing with large numbers of structures, they 
can miss a good answer that is only predicted by a few methods. At least in this case, the confinement
method is able to capture this structure that was otherwise missed.

@@

We have described a computational method called confinement for computing the difference free energy between two
conformational ensembles. We show that the difference free energy can be calculated on ~100 residue sized proteins, even
for large conformational changes. We have demonstrated that it can discriminate the folding preferences of a series of
chameleon proteins. We show that the confinement method can discriminate between the native structure and structure
predictions and can identify the best prediction reliably. Perhaps most importantly, we have shown that it can be used
to give residue-level insights into what are the dominant structural factors in a protein that are responsible for the
difference free energies. The confinement should be useful for protein design, structure prediction, and understanding
the mechanism of conformational change.

A key advantage is that this method does not require any reaction coordinate, or sampling a
pathway from conformation $A$ to $B$.

@@

The confinement method has been described in details by Tyka et al. in reference \cite{Tyka2006} and
Cecchini et  al. \cite{Cecchini2009}. The basic approach of the confinement approach is the same in
both these papers. A thermodynamic cycle is used to compute the free energy between conformations A
and B. However, there are some small technical differences between the two approaches. Here we briefly
describe the procedure that we used.

@@

One advantage of confinement method is that
none of the simulations during the restraining step depends on each other. Therefore, with available computer resources it can be fast 
to compute. The computational cost for a single 20 ns simulation (along the confinement step) of $56$-residue protein 
is only $4$ hours in a single gpu computer  \Alberto{all the different confinements in one GPU?}. All calculations were
performed with the Amber 11 suit of programs~\cite{Case2012,Goetz2012} in combination with the ff99SB
forcefield~\cite{Hornak2006} and the generalized born implicit solvent \cite{Mongan2006}.

>>>>

Understanding the free energy differences between different different conformations of a protein is important for
understanding mechanism in biochemistry. The calculation of such conformational free energy differences is difficult and
computational demanding. In this work, we explore the use of ``the confinement method'' to calculate free energy
differences between protein conformations. This method provides two main advantages: (1) it does not require a reaction
coordinate or transition path and (2) it is fast to compute. The free energy difference can be approximately composed
into per-residue contributions, which provides insight into the physical basis for the preference of one conformation
over another. We show that the confinement method can correctly predict the conformational preferences of chameleon
sequences (nearly identical sequences with different structures), can identify the most native-like structures produced
by structure prediction algorithms, and can yield detailed insight into the mechanisms behind these preferences.

@@

In some problems of protein science, we want to know the relative stability of a protein's conformation $A$ compared to
another of its conformations $B$. We call this the \emph{difference free energy}. For example, in allosteric mechanisms,
a protein often adopts one conformation when a ligand is bound and another conformation when no ligand is
bound~\cite{Elber2007}. To understand the mechanism requires knowledge of the relative free energies of the different
protein conformations.  Another example is in protein structure prediction, such as is in the Critical Assessment of
Structure Prediction (CASP) competition~\cite{Moult2011}. If you are using some computational model and have predicted
two putative native structures, $A$ and $B$, you want to compute which has the lower free energy, in order to know which
is the more native-like.  In a third example, you may want to know whether mutating a few amino acids in a protein could
cause the protein to switch from one stable conformation to another, because that could have important consequences for
biological mechanisms and disease. Fourth, sometimes binding a ligand induces a protein from one conformation to
another. A quantitative understanding of such \emph{induced-fit} situations requires knowledge of the difference free
energy of the protein in the two states. In all these cases, it would be useful to have a computational method that can
be used efficiently with atomically detailed physical forcefield models to compute the free energy differences between
different given protein conformations.

One widely explored strategy is to use molecular dynamics simulations along some putative reaction coordinate pathway
from conformation $A$ to
$B$~\cite{Elber2007,West2007,Haas2007,Jonsson1998,E2007,Dellago2002,Cheng2006,Elber2005,Chipot2007}. The free energy
along this reaction coordinate can then be determined using methods such as umbrella sampling
\cite{Torrie1977,Mascarenhas2013} and the weighted histogram analysis method (WHAM) \cite{Kumar1992}. Such approaches
have several limitations. First, it is necessary to know an efficient reaction pathway from $A$ to $B$. If conformations
$A$ and $B$ are quite different, then it can be challenging to find such paths. Second, these methods are
computationally slow. To get an accurate estimate of the total free energy difference $\Delta G = G_B - G_A$ requires
accurate determinations of the many small free energy differences $A \rightarrow 1 \rightarrow 2 \ldots \rightarrow B$,
and each step requires substantial amounts of sampling.  Third, large errors can accumlate, because the pathway error is
a sum of many errors along the many steps.

The calculation of protein conformational free energy has been successfully attempted by a number of groups
\cite{Ytreber2006a,Shell2010,Ytreberg2006,Zheng2008,Spichty2010,Strajbl2000,Park2008,Tyka2006,Cecchini2009,Ovchinnikov2013}.
Some of these methods, like the reference system method~\cite{Ytreberg2006}, deactivated morphing~\cite{Park2008}, and
the confinement method~\cite{Tyka2006,Cecchini2009,Ovchinnikov2013}, take an alternative strategy for computing
difference free energies that does not require knowing a pathway from $A$ to $B$.

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=3.5 in]{method.pdf}
    \end{center}
    \caption{Graphical representation of the thermodynamic cycle employed in the confinement method.}
\label{fig:method}
\end{figure}

@@

First, we verified that our implementation of the confinement method produces results compatible with previous
calculations reported in the literature.  The method has previously been applied to a 16 amino acid residue
$\beta$-hairpin from protein G, known as BHP~\cite{Cecchini2009}. We calculated the free energy difference between the
native conformation (called bhp1, which has a two-stranded $\beta$-sheet) and a non-native conformation (called bhp3,
which has a three-stranded $\beta$-sheet). Our confinement calculation shows that bhp1 is more stable by 1.7 kcal/mol,
which is consistent with 200 $\mu$s equilibrium simulations showing that bhp1 is more favorable configuration by 1.8
kcal/mol, and in agreement with previous calculations~\cite{Cecchini2009}.

Second, we looked at 6 target proteins from the CASP9 experiment~\cite{Moult2011}. For each target, we examined up to 5
submitted models.  We computed the difference free energy between the experimental native structure and the best model.
Figure~\ref{fig:summary_casp} and Table~\ref{table:casp_control} show that, in 5 out of 6 cases, the confinement method
assigns a lower free energy to the experimentally determined structure than to any of the decoys. Other well-known
discriminators can also successfully tell native structures from computer generated models~\cite{Sheffler2009,Zhou2002};
it is just an independent useful validation that the confinement calculations make sense.

\begin{table}
\begin{center}
\caption{The confinement method assigns a more favorable free energy to the experimentally determined structure than to
    computer-generated predictions. For each target, we examined as many as five predictions submitted by CASP
    participants. We report the free energy difference between the most favorable decoy and the experimentally
    determined structure. Positive $\Delta\Delta G (=\Delta G_{best~decoy} - \Delta G_{native})$ values indicate that
    the experimental structure is predicted to be more favorable than any of the decoys.}
\label{table:casp_control}
\begin{tabular}{l l l l}\hline
    CASP Target  & PDB Identifier & $\Delta \Delta G$ (kcal/mol) & GDT-TS \\ \hline
     T0531       &    2KJX        &          $11.15 \pm 0.70$    &  $44$ \\ \hline
     T0538       &    2L09        &          $-3.00 \pm 0.47$    &  $96$ \\ \hline
     T0540       &    3MX7        &          $16.94 \pm 0.49$    &  $70$ \\ \hline
     T0559       &    2L01        &          $2.10 \pm 0.24$     &  $94$ \\ \hline
     T0560       &    2L02        &          $18.00 \pm 0.49$    &  $94$ \\ \hline
     T0569       &    2KYW        &          $20.01 \pm 0.69$    &  $78$ \\ \hline
\end{tabular}
\end{center}
\end{table}

@@

We also tested the confinement method predictions for difference free energies on the chameleon sequences of Orban et
al.\cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012}. These are instances in which two highly similar sequences
fold into remarkably different structures. Orban and co-workers have designed a protein-G-like sequence of 56-residues
that is marginally stable in one of two possible folds. By mutating key residues in this sequence they are able to
stabilize one fold or the other (see Figure~\ref{fig:orban}). We refer to the $4\beta + \alpha$ structure as the $\beta$
conformation, and we refer to the $3\alpha$ structure as the $\alpha$ conformation. We denote sequences that prefer the
$\alpha$ fold as GA and sequences that prefer $\beta$ as GB. One pair of sequences (GA88/GB88) is 88 percent identical
in sequence, differing in seven positions. Another pair (GA95/GB95) is 95 percent identical, differing in three
positions. Accurately predicting the structural preferences such similar sequences poses a difficult challenge for
computational methods \cite{Allison2011}.

@@

We initially approached this problem by making a model of each sequence with the same backbone structure as its partner
chameleon sequence. For example, we took the sequence of GA88 and built a model with the same overall structure as GB88.
We then used the confinement method to assess the free energy difference between these two structures. In other words,
we are comparing the experimentally determined GA88/$\alpha$ structure with the computer-generated GA88/$\beta$
structure. The confinement method was able to predict the conformational preferences correctly for all four sequences
(data not shown). This is, however, not surprising as it is often easy to distinguish computational models from native
structures~\cite{Handl2009,Sheffler2009}. To avoid this potential problem, we instead computed relative free energy of
two different computer generated models for each sequence. One model is based on the $\alpha$ structure and the other on
the $\beta$ structure (see Supporting Information for details on the modeling procedure). This is a much more realistic
test of the confinement method's ability to accurately calculate the difference free energy.

@@

To better understand the mechanism behind these chameleon proteins, we approximately decomposed~\cite{Tyka2006} the free
energy into per-residue contributions. We decompose each confinement step ($\Delta G_{A,A^\ast}$, $\Delta G_{B,B^\ast}$)
into per-residue contributions. We can also decompose the remaining enthalpy in the confined ensemble into per-residue
contributions. However, the method is approximate because we do not include the residual conformational entropy from the
normal mode or quasiharmonic analysis steps. Computing per-residue contributions helps us to identify important residues
that stabilize a particular conformation. The per-residue difference free energy, $\Delta \Delta G (\beta - \alpha)$, is
shown in Figure~\ref{fig:perresidue_orban}.

@@

Although the overall free energy difference between the two structures is small (within 3--5 kcal/mol), individual
residues can show marked preferences for being in either $\alpha$ or $\beta$ conformation. Such differences can be
understood by looking in detail at the local environment for those residues. For example, the region around residue 7
forms a random coil in the $\alpha$ structure, whereas it forms a beta-sheet structure in the $\beta$ structure. These
residues strongly favour the locally well-packed and hydrogen bonded environment found in the $\beta$ sheet, but in the
alpha conformation there are no other residues to form a beta sheet with. The global conformation of the protein has
``overridden'' the local preference of these residues. Overall, favoring the $\alpha$ or $\beta$ structure is a delicate
balance, where the global free energy difference is small and the contributions of different per-residue tendencies
largely cancel out. For these chameleon proteins, it is therefore very likely that by changing key residue the global
fold preferences can be changed.

Three residues are mutated between GA95 and GB95, at positions 20, 30, and 45. In GA95, these residues (L20, I30 and
L45) stabilize the $\alpha$ structure (Figure~\ref{fig:perresidue_orban}, upper panel). In GB95 (lower panel), two of
these residues (F30 and Y45) favor the $\beta$ structure, because they have large solvent exposed surface areas in the
$\alpha$ structure, but are more buried in the $\beta$ structure. Additionally, Y45 forms a hydrogen bond with D47 in
the $\beta$ structure. On the other hand, residue A20 from GB95 still favors the $\alpha$ structure, although less
strongly than L20 in GA95.

The GA and GB sequences are nearly identical, so there are some common features observed for all sequences.
The experimental observations classified the protein into two parts: Amino acids 9--51 are fully structured in both
folds, whereas residues 1--8 and 52--56 are unstructured in the $\alpha$ fold, but form $\beta$-strands in the $\beta$
fold. Most of the amino acid residues in the region 1--9 have negative per-residue free energies, which means that these
residues favor the $\beta$ structure. Roles of some other important residues in stabilizing either the $\alpha$ or
$\beta$ conformation are summarized in Figure S1.

@@

These per-residue free energy decompositions provide a great deal of insight into the driving forces behind protein
folding and conformational change. We believe that such calculations may also be useful for designing proteins with
specific structures and functions.

@@

We have performed two experiments centered on CASP. In the first experiment, we tried to rank-order predictions for
several targets. For each target, the predictions were either produced by a single group---presumably using the same
method for each prediction, or were produced by several different groups---using different methods. The goal of this
experiment is to determine if the physics-based confinement method can correctly identify more native-like structures as
having lower free energies. Our second experiment was to see if the confinement method can identify structures that are
missed by other meta-prediction servers. Most successful meta-prediction servers are based on the idea of consensus: if
many different prediction methods produce similar results, then that is probably a correct prediction
\cite{Kryshtafovych2011,Wang2011}. This is often a powerful heuristic, but it can miss cases where there is a very good
result that is only predicted by one method. We chose several cases where such structures were missed by the best
meta-predictors in CASP and assessed if the confinement method can correctly identify these accurate models. All of our
calculations were performed after the CASP experiment; they are not blind predictions.

As is common in the CASP experiment, we assess our results in terms of Global Distance Test Total Score (GDT-TS)
\cite{Zemla2003}, which is a C$\alpha$ based measure of structural accuracy. It can be understood roughly as the
percentage of residues that are correctly positioned in the model (range 0 to 100, higher is better). We did not have
enough computer power to analyze every model, so we chose a small subset of examples from a selection of different
server groups that have done well in past CASP events.

@@

\begin{table}
\begin{center}
\caption{The six different CASP targets that are used in this study. The corresponding PDB Identifier and description
    are also listed.}
\label{table:casp_testcase}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier &  Description \\ \hline
     T0531       &    2KJX        &  extracellular domain of the jumping  \\
                 &                &  translocation breakpoint protein  \\ \hline
     T0538       &    2L09        &  protein asr4154 from Nostoc sp. PCC7120     \\ \hline
     T0540       &    3MX7        &  human Fas apoptotic inhibitory molecule     \\ \hline
     T0559       &    2L01        &  protein BVU3908 from Bacteroides vulgatus   \\ \hline
     T0560       &    2L02        &  protein BT2368 from Bacteroides  \\
                 &                &  thetaiotaomicron         \\ \hline
     T0569       &    2KYW        &  domain of adhesion exoprotein from \\
                 &                &   Pediococcus pentosaceus \\ \hline
\end{tabular}
\end{center}
\end{table}

@@

We performed a similar calculation for target T0560 with two models from the group ``Splicer''. The remaining three
models were discarded as they were too similar to the rest of the models. Again, we can correctly identify the native
sate and our calculated free energy based ranking matches well with GDT-TS (Figure S2 in the supporting information).

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=3.5 in]{T0540.pdf}
    \end{center}
    \caption{The confinement method can correctly rank the native structure and two models submitted by different
        prediction groups.}
\label{fig:T0540}
\end{figure}

\subsection*{The per-residue free energy is sensitive to small changes in protein conformation}

In a previous section, we discussed how the per-residue free energy can reveal mechanistic detail behind the
conformational preference of chameleon sequences. In this section, our aim is to apply the same method and try to
understand if the per-residue free energy can help us identify problematic residues in structure predictions. We chose
CASP target T0569 and compared the experimental NMR structure with the best predicted model (GDT-TS=78; predicted by the
``Mufold'' group). The confinement method predicts that the experimental structure is more stable by 20 kcal/mol (Figure
S2).

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=3.5in]{T0569_perres1.pdf}
    \end{center}
    \caption{Per-residue difference free energy between the experimental NMR structure and the best prediction for CASP
        target T0569. The amino acid residues that are colored in deep red and deep blue stabilizes the NMR structure
        and the prediction, respectively; the residues with light blue color do not have a strong preference.}
\label{fig:T0569_per_residue}
\end{figure}

\subsection*{The confinement method occasionally produces incorrect results}

@@

\begin{figure}
\begin{center}
\includegraphics[width=4.9 in,height=4.0 in]{Target_538_compare.pdf}
\end{center}
\caption{Small differences in sidechain orientation can cause large differences in free energy. We compare the
    experimental structure and a computer generated model for CASP target T0538. (A) Two salt bridges between residues
    E35-R32 and E28-K24 exist in the native structure. These are absent in the computer generated model structure.
    Instead, (B) a new salt bridge is formed between residues R32-E28 in model. (C) Another salt bridge between residues
    R26-E50 in the experimental structure. (D) In comparison to the native structure, orientation of F51 differs in the
    model.}
\label{fig:T0538compare}
\end{figure}

@@

A part of the CASP experiment is dedicated to the quality assessment (QA) of predicted models \cite{Kryshtafovych2011}.
Here, predictors were asked to produce an overall score (called QMODE1) for each model on a scale from 0 to 1, with
higher values corresponding to better models \cite{Kryshtafovych2011}. Many of the groups use consensus strategies in
such experiments. Here, we chose two computer generated models from CASP target T0538, where the top performing group
``MUFOLD-WQA'' \cite{Wang2011} failed to identify the best model. We examined two models, one from ``PconsR'' (GDT-TS =
96), which MUFOLD-WQA gave a QMODE1 = 0.54. The second model was from ``MULTICOM-NOVEL'' (GDT-TS=83), which is a poorer
model, but MUFOLD-WQA assigned a higher QMODE1 = 0.59. The confinement method gives a difference free energy that favors
the PconsR model by 3.9 kcal/mol, which correctly identifies the more accurate model. Although consensus prediction
strategies are very effective, they can miss a good prediction that is only predicted by a few methods. At least in this
case, the confinement method is able to capture this structure that was otherwise missed.

@@

We have described a computational method called the confinement method for computing the difference free energy between
two conformational ensembles. We show that the difference free energy can be calculated on ~100 residue sized proteins,
even for large conformational changes. We have demonstrated that it can discriminate the folding preferences of a series
of chameleon proteins. We show that the confinement method can discriminate between the native structure and structure
predictions and can identify the best prediction reliably. Perhaps most importantly, we have shown that it can be used
to give residue-level insights into what are the dominant structural factors in a protein that are responsible for the
difference free energies. The confinement should be useful for protein design, structure prediction, and understanding
the mechanism of conformational change.

@@

The confinement method has been described in detail by Tyka et al. in reference \cite{Tyka2006} and Cecchini et al.
\cite{Cecchini2009}. The basic approach of the confinement approach is the same in both these papers. A thermodynamic
cycle is used to compute the free energy between conformations A and B. However, there are some small technical
differences between the two approaches. Here, we briefly describe the procedure that we used.

@@

One advantage of confinement method is that none of the simulations during the restraining step depends on each other.
Therefore, with available computer resources it can be fast to compute. We ran each confinement calculation on a single
graphics processing unit (GPU). For a 56-residue protein, this leads to a calculation time of only 4 hours on 40 GPUs (1
GPU per confinement calculation x 20 calculations per structure x 2 structures). All calculations were performed with
the Amber 11 suit of programs~\cite{Case2012,Goetz2012} in combination with the ff99SB forcefield~\cite{Hornak2006} and
the GBneck generalized born implicit solvent model \cite{Mongan2006}.
